\vspace{30mm} {\Large\bf RISC-V Duo-PLIC for the Advancced Interrupt Architecture} \\ {\large Document Version \DuoPLICRev}

### 2023-05-09 doc/src/VSLevel.tex

As introduced in Chapter~\ref{ch:CSRs}, several hypervisor and VS CSRs
are added:  \z{hvien}, \z{hvictl}, \z{hviprio1}, \z{hviprio2},
\z{vsiselect}, \z{vsireg}, \z{vstopei}, and \z{vstopi}.
For RV32, the following ``high-half'' CSR partners are also added:
\z{hidelegh}, \z{hvienh}, \z{hviph}, \z{hviprio1h}, \z{hviprio2h},
\z{vsiph}, and \z{vsieh}.
As always, when executing in \mbox{VS-mode} or \mbox{VU-mode}, the
VS CSRs substitute for the corresponding supervisor CSRs.

@@

Instead of a physical VS-level \z{iprio} array, a separate hardware
mechanism is provided for configuring the priorities of a subset
of interrupts for VS level, using hypervisor CSRs \z{hviprio1} and
\z{hviprio2}, plus, for RV32, their high-half partners, \z{hviprio1h}
and \z{hviprio2h}.
The subset of major interrupt numbers whose priority may be configured
in hardware are these:
\begin{displayLinesTable}[c@{\quad}l]
\ 1    & Supervisor software interrupt \\
\ 5    & Supervisor timer interrupt \\
 13    & Counter overflow interrupt \\
14--23 & \em Reserved for standard local interrupts \\
\end{displayLinesTable}
For interrupts directed to VS level, software-configurable priorities
are not supported in hardware for standard local interrupts
in the range 32--48.

@@

For RV32, registers \z{hviprio1}, \z{hviprio1h}, \z{hviprio2}, and
\z{hviprio2h} have these formats:

RV32 \z{hviprio1}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & \em Reserved for priority number for interrupt 0; reads as zero \\
bits 15:8  & Priority number for interrupt 1, supervisor software interrupt \\
bits 23:16 & \em Reserved for priority number for interrupt 4; reads as zero \\
bits 31:24 & Priority number for interrupt 5, supervisor timer interrupt \\
\end{displayLinesTable}
RV32 \z{hviprio1h}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & \em Reserved for priority number for interrupt 8; reads as zero \\
bits 15:8  & Priority number for interrupt 13, counter overflow interrupt \\
bits 23:16 & Priority number for interrupt 14 \\
bits 31:24 & Priority number for interrupt 15 \\
\end{displayLinesTable}
RV32 \z{hviprio2}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 16 \\
bits 15:8  & Priority number for interrupt 17 \\
bits 23:16 & Priority number for interrupt 18 \\
bits 31:24 & Priority number for interrupt 19 \\
\end{displayLinesTable}
RV32 \z{hviprio2h}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 20 \\
bits 15:8  & Priority number for interrupt 21 \\
bits 23:16 & Priority number for interrupt 22 \\
bits 31:24 & Priority number for interrupt 23 \\
\end{displayLinesTable}

For RV64, registers \z{hviprio1} and \z{hviprio2} are the \mbox{64-bit}
concatenations of the same RV32 registers with their respective
high-half partners:

RV64 \z{hviprio1}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & \em Reserved for priority number for interrupt 0; reads as zero \\
bits 15:8  & Priority number for interrupt 1, supervisor software interrupt \\
bits 23:16 & \em Reserved for priority number for interrupt 4; reads as zero \\
bits 31:24 & Priority number for interrupt 5, supervisor timer interrupt \\
bits 39:32 & \em Reserved for priority number for interrupt 8; reads as zero \\
bits 47:40 & Priority number for interrupt 13, counter overflow interrupt \\
bits 55:48 & Priority number for interrupt 14 \\
bits 63:56 & Priority number for interrupt 15 \\
\end{displayLinesTable}
RV64 \z{hviprio2}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 16 \\
bits 15:8  & Priority number for interrupt 17 \\
bits 23:16 & Priority number for interrupt 18 \\
bits 31:24 & Priority number for interrupt 19 \\
bits 39:32 & Priority number for interrupt 20 \\
bits 47:40 & Priority number for interrupt 21 \\
bits 55:48 & Priority number for interrupt 22 \\
bits 63:56 & Priority number for interrupt 23 \\
\end{displayLinesTable}

Each priority number in \z{hviprio1}/\z{hviprio1h} and
\z{hviprio2}/\z{hviprio2h} is a {\WARL} unsigned integer field that
is either read-only zero or implements a minimum of IPRIOLEN bits or
6~bits,
whichever is larger, and preferably all 8~bits.
Implementations may freely choose which priority number fields are
read-only zeros, but all other fields must implement the same number of
integer bits.
A minimal implementation of these CSRs has them all be read-only zeros.

A hypervisor can choose to employ registers \z{hviprio1}/\z{hviprio1h}
and \z{hviprio2}/\z{hviprio2h} when emulating the (virtual)
supervisor-level \z{iprio} array accessed indirectly through
\z{siselect} and \z{sireg} (really \z{vsiselect} and \z{vsireg}) for a
virtual hart.
For interrupts not in the subset supported by
\z{hviprio1}/\z{hviprio1h} and \z{hviprio2}/\z{hviprio2h}, the priority
number bytes in the emulated \z{iprio} array can be read-only zeros.

@@

If a hypervisor really must emulate configurability of priority for
interrupts beyond the subset supported by \z{hviprio1}/\z{hviprio1h}
and \z{hviprio2}/\z{hviprio2h}, it can do so with extra effort
by setting bit VTI of CSR \z{hvictl}, described in the next
subsection.

@@

Assuming a virtual hart does not need configurable priority for
major interrupts beyond the subset supported in hardware by
\z{hviprio1}/\z{hviprio1h} and \z{hviprio2}/\z{hviprio2h}, a hypervisor
can assert interrupts to the virtual hart using CSRs \z{hvien}
(Hypervisor Virtual-Interrupt-Enable) and \z{hvip} (Hypervisor
Virtual-Interrupt-Pending bits), plus their high-half partners for RV32,
\z{hvienh} and \z{hviph}.
These CSRs affect interrupts for VS level much the same way that
\z{mvien}/\z{mvienh} and \z{mvip}/\z{mviph} do for supervisor level, as
explained in Section~\ref{sec:virtIntrs-S}.

Each bit of registers \z{hvien}/\z{hvienh} and \z{hvip}/\z{hviph}
corresponds with an interrupt number in the range 0--63.
Bits 12:0 of \z{hvien} are reserved and must be read-only zeros,
while bits 12:0 of \z{hvip} are defined by the {\RISCV} Privileged
Architecture.
Specifically, bits 10, 6, and 2 of \z{hvip} are writable bits that
correspond to VS-level external interrupts (VSEIP), VS-level timer
interrupts (VSTIP), and VS-level software interrupts (VSSIP),
respectively.

The following applies only to the CSR bits for interrupt numbers 13--63:
When a bit in \z{hideleg}/\z{hidelegh} is one, then the same bit
position in \z{vsip}/\z{vsiph} is an alias for the corresponding bit in
\z{sip}/\z{siph}.
Else, when a bit in \z{hideleg}/\z{hidelegh} is zero and the
matching bit in \z{hvien}/\z{hvienh} is one, the same bit position
in \z{vsip}/\z{vsiph} is an alias for the corresponding bit in
\z{hvip}/\z{hviph}.
A bit in \z{vsip}/\z{vsiph} is read-only zero when the corresponding
bits in \z{hideleg}/\z{hidelegh} and \z{hvien}/\z{hvienh} are both
zero.
The combined effects of \z{hideleg}/\z{hidelegh} and
\z{hvien}/\z{hvienh} on \z{vsip}/\z{vsiph} and \z{vsie}/\z{vsieh} are
summarized in Table~\ref{tab:intrFilteringForVS}.

@@

For interrupt numbers 13--63, a bit in \z{vsie}/\z{vsieh} is
writable if and only if the corresponding bit is set in either
\z{hideleg}/\z{hidelegh} or \z{hvien}/\z{hvienh}.
When an interrupt is delegated by \z{hideleg}/\z{hidelegh}, the
writable bit in \z{vsie}/\z{vsieh} is an alias of the corresponding bit
in \z{sie}/\z{sieh};
else it is an independent writable bit.
The Privileged Architecture specifies when bits 12:0 of \z{vsie} are
aliases of bits in \z{hie}.
As usual, bits that are not writable in \z{vsie}/\z{vsieh} must be
read-only zeros.

If a bit of \z{hideleg}/\z{hidelegh} is zero and the corresponding bit
in \z{hvien}/\z{hvienh} is changed from zero to one, then the value
of the matching bit in \z{vsie}/\z{vsieh} becomes {\unspecified}.
Likewise, if a bit of \z{hvien}/\z{hvienh} is one and the corresponding
bit in \z{hideleg}/\z{hidelegh} is changed from one to zero, the value
of the matching bit in \z{vsie}/\z{vsieh} again becomes {\unspecified}.

For interrupt numbers 13--63, implementations may freely choose which
bits of \z{hvien}/\z{hvienh} are writable and which bits are read-only
zero or one.
If such a bit in \z{hvien}/\z{hvienh} is read-only zero (preventing the
virtual interrupt from being enabled), the same bit should be read-only
zero in \z{hvip}/\z{hviph}.
All other bits for interrupts 13--63 must be writable in
\z{hvip}/\z{hviph}.

@@

\item
asserting for VS level a major interrupt not supported by
\z{hvien}/\z{hvienh} and \z{hvip}/\z{hviph};

\item
implementing configurability of priorities at VS level for major
interrupts beyond those supported by \z{hviprio1}/\z{hviprio1h} and
\z{hviprio2}/\z{hviprio2}; or

@@

When bit VTI (Virtual Trap Interrupt control) =~1, attempts from
\mbox{VS-mode} to explicitly access CSRs \z{sip}, \z{sie}, or, for RV32
only, \z{siph} or \z{sieh}, cause a virtual instruction exception.
Furthermore, for any given CSR, if there is some circumstance in which
a write to the register may cause a bit of \z{vsip}/\z{vsiph} to change
from one to zero, excluding bit~9 for external interrupts (SEIP),
then when VTI =~1, a virtual instruction exception is raised
also for any attempt by the guest to write this register.
Both the value being written to the CSR and the value of
\z{vsip}/\z{vsiph} (before or after) are ignored for determining
whether to raise the exception.
(Hence a write would not actually need to change a bit of
\z{vsip}/\z{vsiph} from one to zero for the exception to be raised.)
In particular, if register \z{vstimecmp} is implemented
(from extension Sstc), then attempts from \mbox{VS-mode}
to write to \z{stimecmp} or (RV32 only) \z{stimecmph}
cause a virtual instruction exception when VTI~=~1.

\begin{commentary}
For the standard local interrupts
(major identities 13--23 and 32--47), and for
software interrupts (SSI), the corresponding interrupt-pending
bits in \z{vsip}/\z{vsiph} are defined as ``sticky,''
meaning a guest can clear them only by writing directly
to \z{sip}/\z{siph} (really \z{vsip}/\z{vsiph}).
Among the standard-defined interrupts, that leaves only
timer interrupts (STI), which can potentially be cleared
in \z{vsip} by writing a new value to \z{vstimecmp}.
\end{commentary}

@@

\item
if \z{hvictl}.VTI =~0:
\begin{tightList}
\item[+]
the highest-priority pending-and-enabled major interrupt indicated by
\z{vsip}/\z{vsiph} and \z{vsie}/\z{vsieh} other than a supervisor
external interrupt (code 9), using the priority numbers assigned by
\z{hviprio1}/\z{hviprio1h} and \z{hviprio2}/\z{hviprio2h};
\end{tightList}

@@

The Advanced Interrupt Architecture modifies the standard
{\RISCV} Privileged Architecture such that an interrupt
is pending at VS level if and only if \z{vstopi} is not zero.
CSRs \z{vsiph}/\z{vsip} and \z{vsieh}/\z{vsie} do not by themselves
determine whether a VS-level interrupt is pending, though
they may do so indirectly through their effect on \z{vstopi}.

>>>>

As introduced in Chapter~\ref{ch:CSRs}, several hypervisor and VS CSRs
are added:  \z{hvien}, \z{hvictl}, \z{hviprio1}, \z{hviprio2},
\z{vsiselect}, \z{vsireg}, \z{vstopei}, and \z{vstopi}.
(And for RV32, the following high-half CSRs are also added:
\z{hidelegh}, \z{hvienh}, \z{hviph}, \z{hviprio1h}, \z{hviprio2h},
\z{vsiph}, and \z{vsieh}.)
As always, when executing in \mbox{VS-mode} or \mbox{VU-mode}, the
VS CSRs substitute for the corresponding supervisor CSRs.

@@

Instead of a physical VS-level \z{iprio} array, a separate hardware
mechanism is provided for configuring the priorities of a subset
of interrupts for VS level, using hypervisor CSRs \z{hviprio1} and
\z{hviprio2}.
The subset of major interrupt numbers whose priority may be configured
in hardware are these:
\begin{displayLinesTable}[c@{\quad}l]
\ 1    & Supervisor software interrupt \\
\ 5    & Supervisor timer interrupt \\
 13    & Counter overflow interrupt \\
14--23 & \em Reserved for standard local interrupts \\
\end{displayLinesTable}
For interrupts directed to VS level, software-configurable priorities
are not supported in hardware for standard local interrupts
in the range 32--48.

@@

Registers \z{hviprio1} and \z{hviprio2} have these formats:

\z{hviprio1}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & \em Reserved for priority number for interrupt 0; reads as zero \\
bits 15:8  & Priority number for interrupt 1, supervisor software interrupt \\
bits 23:16 & \em Reserved for priority number for interrupt 4; reads as zero \\
bits 31:24 & Priority number for interrupt 5, supervisor timer interrupt \\
bits 39:32 & \em Reserved for priority number for interrupt 8; reads as zero \\
bits 47:40 & Priority number for interrupt 13, counter overflow interrupt \\
bits 55:48 & Priority number for interrupt 14 \\
bits 63:56 & Priority number for interrupt 15 \\
\end{displayLinesTable}
\z{hviprio2}:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l]
bits 7:0   & Priority number for interrupt 16 \\
bits 15:8  & Priority number for interrupt 17 \\
bits 23:16 & Priority number for interrupt 18 \\
bits 31:24 & Priority number for interrupt 19 \\
bits 39:32 & Priority number for interrupt 20 \\
bits 47:40 & Priority number for interrupt 21 \\
bits 55:48 & Priority number for interrupt 22 \\
bits 63:56 & Priority number for interrupt 23 \\
\end{displayLinesTable}

Each priority number in \z{hviprio1} and
\z{hviprio2} is a {\WARL} unsigned integer field that
is either read-only zero or implements a minimum of IPRIOLEN bits or
6~bits,
whichever is larger, and preferably all 8~bits.
Implementations may freely choose which priority number fields are
read-only zeros, but all other fields must implement the same number of
integer bits.
A minimal implementation of these CSRs has them both be read-only zeros.

A hypervisor can choose to employ registers \z{hviprio1}
and \z{hviprio2} when emulating the (virtual)
supervisor-level \z{iprio} array accessed indirectly through
\z{siselect} and \z{sireg} (really \z{vsiselect} and \z{vsireg}) for a
virtual hart.
For interrupts not in the subset supported by
\z{hviprio1} and \z{hviprio2}, the priority
number bytes in the emulated \z{iprio} array can be read-only zeros.

@@

If a hypervisor really must emulate configurability of priority for
interrupts beyond the subset supported by \z{hviprio1}
and \z{hviprio2}, it can do so with extra effort
by setting bit VTI of CSR \z{hvictl}, described in the next
subsection.

@@

Assuming a virtual hart does not need configurable priority for
major interrupts beyond the subset supported in hardware by
\z{hviprio1} and \z{hviprio2}, a hypervisor
can assert interrupts to the virtual hart using CSRs \z{hvien}
(Hypervisor Virtual-Interrupt-Enable) and \z{hvip} (Hypervisor
Virtual-Interrupt-Pending bits).
These CSRs affect interrupts for VS level much the same way that
\z{mvien} and \z{mvip} do for supervisor level, as
explained in Section~\ref{sec:virtIntrs-S}.

Each bit of registers \z{hvien} and \z{hvip}
corresponds with an interrupt number in the range 0--63.
Bits 12:0 of \z{hvien} are reserved and must be read-only zeros,
while bits 12:0 of \z{hvip} are defined by the {\RISCV} Privileged
Architecture.
Specifically, bits 10, 6, and 2 of \z{hvip} are writable bits that
correspond to VS-level external interrupts (VSEIP), VS-level timer
interrupts (VSTIP), and VS-level software interrupts (VSSIP),
respectively.

The following applies only to the CSR bits for interrupt numbers 13--63:
When a bit in \z{hideleg} is one, then the same bit
position in \z{vsip} is an alias for the corresponding bit in
\z{sip}.
Else, when a bit in \z{hideleg} is zero and the
matching bit in \z{hvien} is one, the same bit position
in \z{vsip} is an alias for the corresponding bit in
\z{hvip}.
A bit in \z{vsip} is read-only zero when the corresponding
bits in \z{hideleg} and \z{hvien} are both
zero.
The combined effects of \z{hideleg} and
\z{hvien} on \z{vsip} and \z{vsie} are
summarized in Table~\ref{tab:intrFilteringForVS}.

@@

For interrupt numbers 13--63, a bit in \z{vsie} is
writable if and only if the corresponding bit is set in either
\z{hideleg} or \z{hvien}.
When an interrupt is delegated by \z{hideleg}, the
writable bit in \z{vsie} is an alias of the corresponding bit
in \z{sie};
else it is an independent writable bit.
The Privileged Architecture specifies when bits 12:0 of \z{vsie} are
aliases of bits in \z{hie}.
As usual, bits that are not writable in \z{vsie} must be
read-only zeros.

If a bit of \z{hideleg} is zero and the corresponding bit
in \z{hvien} is changed from zero to one, then the value
of the matching bit in \z{vsie} becomes {\unspecified}.
Likewise, if a bit of \z{hvien} is one and the corresponding
bit in \z{hideleg} is changed from one to zero, the value
of the matching bit in \z{vsie} again becomes {\unspecified}.

For interrupt numbers 13--63, implementations may freely choose which
bits of \z{hvien} are writable and which bits are read-only
zero or one.
If such a bit in \z{hvien} is read-only zero (preventing the
virtual interrupt from being enabled), the same bit should be read-only
zero in \z{hvip}.
All other bits for interrupts 13--63 must be writable in
\z{hvip}.

@@

\item
asserting for VS level a major interrupt not supported by
\z{hvien} and \z{hvip};

\item
implementing configurability of priorities at VS level for major
interrupts beyond those supported by \z{hviprio1} and
\z{hviprio2}; or

@@

When bit VTI (Virtual Trap Interrupt control) =~1, attempts from
\mbox{VS-mode} to explicitly access CSRs \z{sip} and \z{sie} (or, for RV32
only, \z{siph} and \z{sieh}) cause a virtual instruction exception.
Furthermore, for any given CSR, if there is some circumstance in which
a write to the register may cause a bit of \z{vsip} to change
from one to zero, excluding bit~9 for external interrupts (SEIP),
then when VTI =~1, a virtual instruction exception is raised
also for any attempt by the guest to write this register.
Both the value being written to the CSR and the value of
\z{vsip} (before or after) are ignored for determining
whether to raise the exception.
(Hence a write would not actually need to change a bit of
\z{vsip} from one to zero for the exception to be raised.)
In particular, if register \z{vstimecmp} is implemented
(from extension Sstc), then attempts from \mbox{VS-mode}
to write to \z{stimecmp} (or, for RV32 only, \z{stimecmph})
cause a virtual instruction exception when VTI~=~1.

\begin{commentary}
For the standard local interrupts
(major identities 13--23 and 32--47), and for
software interrupts (SSI), the corresponding interrupt-pending
bits in \z{vsip} are defined as ``sticky,''
meaning a guest can clear them only by writing directly
to \z{sip} (really \z{vsip}).
Among the standard-defined interrupts, that leaves only
timer interrupts (STI), which can potentially be cleared
in \z{vsip} by writing a new value to \z{vstimecmp}.
\end{commentary}

@@

\item
if \z{hvictl}.VTI =~0:
\begin{tightList}
\item[+]
the highest-priority pending-and-enabled major interrupt indicated by
\z{vsip} and \z{vsie} other than a supervisor
external interrupt (code 9), using the priority numbers assigned by
\z{hviprio1} and \z{hviprio2};
\end{tightList}

@@

The Advanced Interrupt Architecture modifies the standard
{\RISCV} Privileged Architecture such that an interrupt
is pending at VS level if and only if \z{vstopi} is not zero.
CSRs \z{vsip} and \z{vsie} do not by themselves
determine whether a VS-level interrupt is pending, though
they may do so indirectly through their effect on \z{vstopi}.
